---
title: "Estudo Livro R for Data Science - Cap 5"
author: "Bruno de A. Machado"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
       highlight: textmate
       theme: flatly
       number_sections: yes
       toc: yes
       toc_float:
         collapsed: yes
         smooth_scroll: no
--- 

# Workflow : Data Transformation 

## Intro

Neste Capítulo do livro [R for Data Science](https://r4ds.had.co.nz/index.html) [^1] é estudado como transformar os dados utilizando `dplyr`, pacote que faz parte do ambiente **tidyverse** .

```{r , message=FALSE, warning=FALSE}
library(tidyverse)
library(nycflights13)
```

Explorando `nycflights13`

```{r}
flights
```

```{r}
glimpse(flights)
```

## dplyr Basics

**5 Key functions**

1. `filter()` : filtrar observações por valores
2. `arrange()`: re-ordernar linhas
3. `select()` : selecionar observações por nomes
4. `mutate()` : criar novas observações 
5. `summarise()` : fornece um resumo estatístico

Todas essas cinco funções podem ser utilizadas em conjunto com `groupby()`. 


## filter()

Com a funções `filter()` podemos criar subsets do conjunto de dados 

```{r}
(dec25 <- filter(flights, month == 12, day == 25))

```
*Dica : * utilizando parantereses, podemos criar um novo dataframe e ao mesmo tempo imprimir a saída.

### Exercícios {.tabset .tabset-fade}

#### 

  **Find all flights that**

Had an arrival delay of two or more hours

```{r}
filter(flights, arr_delay >= 2)
```



Flew to Houston (IAH or HOU)

```{r}

#Op1: usando in
#filter(flights, dest %in% c("IAH","HOU"))

#Op2: usando or
filter(flights, dest == 'IAH' |  dest == 'HOU')
```




Were operated by United, American, or Delta

Verificando os códigos
```{r}
airlines
```

```{r}
filter(flights, carrier %in% c('AA', 'DL', 'UA'))
```




Departed in summer (July, August, and September)

```{r}

#Op1 : usando in
#filter(flights, month %in% 7:9)) 
#filter(flights, month %in% c(7,8,9)) 

#Op2 : usando and
filter(flights, month > 6 & month < 10 )
```


Arrived more than two hours late, but didn’t leave late

```{r}
filter(flights, arr_delay > 120 & dep_delay <= 0)
```



Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
filter(flights, dep_delay >= 60, dep_delay - arr_delay > 30 )
```



Departed between midnight and 6am (inclusive)

```{r}
filter(flights, dep_time <= 600 | dep_time == 2400)
```



####

  **Pergunta** : Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?



```{r}
filter(flights, between(month, 7, 9))
```



####

  **Pergunta** How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

```{r}
filter(flights, is.na(dep_time))
```

####

  **Pergunta** Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)



## arrange()

Utilizado para reordernar as colunas do dataframe e modificar os nomes

```{r}
arrange(flights, year, month, day)
```


```{r}
arrange(flights, desc(dep_delay))
```




### Exercícios 5.3.1 {.tabset .tabset-fade}

#### 

  **Pergunta** : How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).

```{r}
#Op1:
#arrange(flights, desc(is.na(dep_time)), dep_time)

#Op2:
arrange(flights, dep_time) %>% tail()
```


#### 

  **Pergunta** : Sort flights to find the most delayed flights. Find the flights that left earliest.

```{r}
arrange(flights, desc(dep_delay))
```

#### 

  **Pergunta** : Sort flights to find the fastest (highest speed) flights.


```{r}
arrange(flights, air_time )
```

#### 

  **Pergunta** : Which flights travelled the farthest? Which travelled the shortest?

Mais distantes : 
```{r}
arrange(flights, distance)
```

Mais rápidos:
```{r}
arrange(flights, desc(air_time))
```



## select()

Função útil para lidar com datasets grandes, possibilitando criar subsets de forma rápida e fácil


*Selecionando do dataset flights as colunas entre year e day*
```{r}
select(flights, year:day)

```


*Utilizando - podemos excluir do dataset as colunas entre year e day*
```{r}
select(flights, -(year:day))

```


Outras funções que podemos utilizar em conjunto :

* starts_with("abc")
* ends_with("xyz")
* contains("ijk")
* matches("(.)\\1")
* num_range("x", 1:3)

* rename() para renomear colunas `rename(flights, tail_num = tailnum)`

```{r}
select(flights, time_hour, air_time, everything())

```


### Exercícios 5.4.1 {.tabset .tabset-fade}

#### 

  **Pergunta** : Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.

```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)

#select(flights, 4, 6, 7, 9)
#select(flights, all_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
#select(flights, any_of(c("dep_time", "dep_delay", "arr_time", "arr_delay")))
#variables <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
#select(flights, all_of(variables))
```



#### 

  **Pergunta** : What happens if you include the name of a variable multiple times in a select() call?
  
```{r}
select(flights, year, month, day, year, year)

```
  


#### 

  **Pergunta** : What does the any_of() function do? Why might it be helpful in conjunction with this vector?
  `vars <- c("year", "month", "day", "dep_delay", "arr_delay")`
  
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")

select(flights, any_of(vars))
```
  
  
#### 

  **Pergunta** : Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?
`select(flights, contains("TIME"))`

```{r}
select(flights, contains("TIME"))
```
by default it ignore case sensitive.



## mutate()













